Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-22 paper/paper.tex
submitted abstract

  Based on these category-theoretic constructions we propose a new
  syntactic formalism for formal grammars: a version of dependent
  linear-non-linear type theory in which the tensor product is not
  commutative. The linear types can be interpreted as grammars, and
  linear terms as parse transformers. The non-dependent fragment of
  this type theory is already enough to express regular/context-free
  expressions as well as finite automata, with the isomorphism between
  regular languages and automata definable as linear terms. This
  generalizes the language theoretic formalisms such as Kleene
  algebras to incorporate ambiguity. By incorporating a dependency of
  linear types on non-linear types, we can further express indexed
  grammars, pushdown automata and Turing machines. This provides a new
  form of ``logical characterization'' of grammar classes based on a
  substructural logic rather than ordinary first-order logic. Further,
  the type theory can not only express the grammars as types, but
  equivalences between grammars and parsers as terms of the grammar.

>>>>

  Based on these category-theoretic constructions we propose a new
  syntactic formalism for formal grammars: a version of dependent
  linear-non-linear type theory in which the tensor product is not
  commutative. The linear types can be interpreted as grammars, and
  linear terms as parse transformers. The non-dependent fragment of
  this type theory is already enough to express regular/context-free
  expressions as well as finite automata, with the isomorphism between
  regular languages and automata definable as linear terms. This
  generalizes the language theoretic formalisms such as Kleene
  algebras to incorporate ambiguity. By incorporating a dependency of
  linear types on non-linear types, we can express richer grammar
  formalisms such as indexed grammars, pushdown automata and Turing
  machines. This provides a new form of ``logical characterization''
  of grammar classes based on a substructural logic rather than
  ordinary first-order logic. Further, the type theory can not only
  express the grammars as types, but equivalences between grammars and
  parsers as terms of the grammar.
